Examples for Interesting Formulas

### 2021-05-11 assertions/examples/main.tex
Updates assertion examples document

% Packages
\usepackage{xcolor}
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref} % for url command
\usepackage{amsmath}
\usepackage{xfrac}
\usepackage{listings}

% Command defines
\newcommand{\narrow}{\setlength\itemsep{0pt}}
\newcounter{todos}
\setcounter{todos}{0}
\newcommand{\todo}[1]{\textbf{\textcolor{red}{(TODO: #1)}}\refstepcounter{todos}\label{todo \thetodos}}

@@

\begin{itemize}\narrow
    \item \textbf{Logic}
    \begin{itemize}
        \item \textcolor{olive}{CNF SAT Problem}
        \item \textcolor{olive}{DNF SAT Problem}
    \end{itemize}
    \item \textbf{Numerical}
    \begin{itemize}
        \item Number is a prime number
        \item One number is the greatest common divisor ($gcd$) of two other numbers
        \begin{itemize}
            \item \textcolor{olive}{Special case: Two numbers are coprime, i.e., $gcd(a,b) = 1$}
        \end{itemize}
        \item One number is the least common multiple of two other numbers
    \end{itemize}
\item \textbf{Arrays}
    \begin{itemize}
        \item Array is sorted
        \item Fulfills (balanced) binary search tree property, i.e., the left elements of subtree are smaller, right elements of subtree are greater than parent
        \begin{itemize}
            \item Problem: handling trees which are not full, e.g. by defining empty elements?!
        \end{itemize}
        \item \textcolor{olive}{Fulfills heap property, i.e.}
        \begin{itemize}
            \item Max-Heap: any child is smaller or equal than its parent
            \item \textcolor{olive}{Min-Heap: any child is greater or equal than its parent}
            \item Problem: handling trees which are not full, e.g. by defining empty elements?!
        \end{itemize}
        \item All Elements in one are bigger / smaller than in another
        \item One array is the reverse of another (similar to sorted array)
    \end{itemize}
    \item \textbf{Strings}
    \begin{itemize}
        \item All characters are upper case, lower case, \textcolor{olive}{alphanumerical}, ...
        \item One string is (not) a substring of another string (using slices)
    \end{itemize}
    \item \textbf{Sets}
    \begin{itemize}
        \item Set property is fulfilled: no element occurs twice (if set is represented as an array)
        \item \textcolor{olive}{Two sets are disjoint}
    \end{itemize}
\end{itemize}

@@

\subsubsection{Formulation in predicate logic}
Given that $F$ and $\mathcal{A}$ are encoded like this, the assumption that $\mathcal{A}$ satisfies $F$ can be expressed in predicate logic as follows:
\begin{gather*}\label{eq:cnf_sat_formal}
\begin{aligned}
(\forall i, \exists j) (0 \leq i < |F|) \wedge (0 \leq j < |F[i]|) &\Rightarrow ((isPos(F[i][j]) \wedge \mathcal{A}[decode(F[i][j])]) \\
&\vee (\neg isPos(F[i][j]) \wedge \neg \mathcal{A}[decode(F[i][j])])))
\end{aligned}
\end{gather*}

The transformed formula to check for any counterexamples is then:
\begin{gather*}\label{eq:cnf_sat_formal}
\begin{aligned}
(\exists i, \forall j) (0 \leq i < |F|) \wedge (0 \leq j < |F[i]|) &\wedge ((isPos(F[i][j]) \wedge \neg \mathcal{A}[decode(F[i][j])]) \\
&\vee (\neg isPos(F[i][j]) \wedge \mathcal{A}[decode(F[i][j])])))
\end{aligned}
\end{gather*}

@@

%----------------------------------------------------------------------------------------
%	COPRIME
%----------------------------------------------------------------------------------------
\subsection{Two numbers are coprime}\label{subsec:coprime}
The greatest common divisor ($gcd$) of two non zero integers is the largest integer that divides both evenly \cite{knuthtaocp02}. A more specific $gcd$-problem is to find two numbers which are coprime, also called relative prime. Coprime expresses the relation between two integers a and b, where the greatest common divisor between those two is one \cite{coprimewiki}, i.e., $gcd(a, b) = 1$. To check if two numbers are coprime it has to be ensured that there is no greater $gcd$ than one, which can be stated in a propositional formula.
\begin{equation}\label{eq:coprime-universial}
    (a < b): (\forall n) (2 \le n \le a) \Rightarrow \neg((a \mathbin{\%} n) = 0 \land (b \mathbin{\%} n) = 0)
\end{equation}
To check this it would take $\mathcal{O}(a)$, means linear time complexity. To find a contradiction to this condition we can formulate:
\begin{equation}\label{eq:coprime-existential}
    (a < b): (\exists n) (2 \le n \le a) \land ((a \mathbin{\%} n) = 0 \land (b \mathbin{\%} n) = 0)
\end{equation}
When it is ensured that the greater number is not evenly divided by the smaller one ($b\mathbin{\%}a=0$), the search space can reduced by half to $(2 \le n \le \lfloor \frac{a}{2} \rfloor)$, because in the upper half of the lower number no integer can evenly divide this number.

@@

\subsubsection{Least common multiple}
A similar problem would be to check the least common multiple ($lcm$) of two numbers. There the space between the greater number and the stated $lcm$ has to be checked for a smaller one. This search space can be significantly reduced by only selecting numbers, which are multiples of the greater number and check if the smaller number can divide this number without remainder.
\begin{equation}\label{eq:lcm-existential}
    (a < b): (\exists n) (2 \le n < \frac{lcm}{b}) \land ((b*n) \mathbin{\%} a = 0)
\end{equation}

@@

\begin{lstlisting}[language=Solidity, style=Solidity, numbers=none, caption=Access a heap in array representation]
heap_array[(k-1)/2] // Get parent of actual node k
heap_array[(k*2)+1] // Get left child of actual node k
heap_array[(k*2)+2] // Get right child of actual node k
\end{lstlisting}
So to check if an array $a$ fulfills the min heap property, we can formulate it in the following predicate logic:
\begin{equation}\label{eq:heap-unversial}
  (\forall k) (1 \le k < |a|) \Rightarrow a[\lfloor(k-1)/2)\rfloor] \le a[k]
\end{equation}
This equation checks the direction from the child to the parent node. (Following the Exposé) The direction could also be turned around and every parent node could be checked against its child nodes, which does not change the number of comparisons.
\begin{equation}\label{eq:heap-unversial}
  (\forall k) (0 \le k \le \lfloor |a|/2 \rfloor) \Rightarrow a[k] \le a[2k+1] \land (2k+2 < |a| \Rightarrow a[k] \le a[2k+2])
\end{equation}
This test would need $\mathcal{O}(|a|)$, i.e., linear time complexity. To find a contradiction this can be formulated to
\begin{equation}\label{eq:heap-unversial}
  (\exists k) (1 \le k \le |a|) \land a[\lfloor(k-1)/2)\rfloor] > a[k]
\end{equation}
This example is similar to the example for checking if an array is sorted. Further such examples could be:
\begin{itemize}
    \item One array is the reverse of another
    \item Array fulfills binary search tree property (If not balanced, empty fields must be identified correctly)
\end{itemize}

>>>>

% Packages
\usepackage{xcolor}
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref} % for url command
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{xfrac}
\usepackage{listings}

% Command defines
\newcommand{\narrow}{\setlength\itemsep{0pt}}
\newcounter{todos}
\setcounter{todos}{0}
\newcommand{\todo}[1]{\textbf{\textcolor{red}{(TODO: #1)}}\refstepcounter{todos}\label{todo \thetodos}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\usepackage{makecell}

@@

\begin{itemize}\narrow
    \item \textbf{Logic}
    \begin{itemize}
        \item \textcolor{olive}{CNF SAT Problem}
        \item \textcolor{olive}{DNF SAT Problem}
    \end{itemize}
    \item \textbf{Numerical}
    \begin{itemize}
        \item Number is a prime number
        \item One number is the greatest common divisor ($gcd$) of two other numbers
        \begin{itemize}
            \item \textcolor{olive}{Special case: Two numbers are coprime, i.e., $gcd(a,b) = 1$}
        \end{itemize}
        \item One number is the least common multiple of two other numbers
    \end{itemize}
    \item \textbf{Arrays}
    \begin{itemize}
        \item Array is sorted
        \item Fulfills (balanced) binary search tree property, i.e., the left elements of subtree are smaller, right elements of subtree are greater than parent
        \begin{itemize}
            \item Problem: handling trees which are not full, e.g. by defining empty elements?!
        \end{itemize}
        \item \textcolor{olive}{Fulfills heap property, i.e.}
        \begin{itemize}
            \item Max-Heap: any child is smaller or equal than its parent
            \item \textcolor{olive}{Min-Heap: any child is greater or equal than its parent}
            \item Problem: handling trees which are not full, e.g. by defining empty elements?!
        \end{itemize}
        \item All Elements in one are bigger / smaller than in another
        \item One array is the reverse of another (similar to sorted array)
    \end{itemize}
    \item \textbf{Strings}
    \begin{itemize}
        \item All characters are upper case, lower case, \textcolor{olive}{alphanumerical}, ...
        \item One string is (not) a substring of another string (using slices)
    \end{itemize}
    \item \textbf{Sets}
    \begin{itemize}
        \item Set property is fulfilled: no element occurs twice (if set is represented as an array)
        \item \textcolor{olive}{Two sets are disjoint}
    \end{itemize}
    \item \textbf{Matrices}
    \begin{itemize}
        \item Identity matrix $\mathbb{I}$: square matrix with only diagonal ones, else zeros
        \item Logical matrix (only one or zero entries), matrix of ones (only one entries), zero matrix, ...
        \item Symmetric matrix: $\mat{A} = \mat{A}^T$
        \item Orthogonal matrix: $\mat{A}\mat{A}^T = \mathbb{I}$
        \item Correct calculation result, e.g. matrix multiplication, multiplication with a constant, addition, ... (which might be more involved and mostly need all terms for verification, which is memory consuming)
    \end{itemize}
\end{itemize}

@@

\subsubsection{Formulation in predicate logic}
Given that $F$ and $\mathcal{A}$ are encoded like this, the assumption that $\mathcal{A}$ satisfies $F$ can be expressed in predicate logic as follows:
\begin{gather*}\label{eq:cnf_sat_formal_1}
\begin{aligned}
(\forall i, \exists j) (0 \leq i < |F|) \wedge (0 \leq j < |F[i]|) &\Rightarrow ((isPos(F[i][j]) \wedge \mathcal{A}[decode(F[i][j])]) \\
&\vee (\neg isPos(F[i][j]) \wedge \neg \mathcal{A}[decode(F[i][j])])))
\end{aligned}
\end{gather*}

The transformed formula to check for any counterexamples is then:
\begin{gather*}\label{eq:cnf_sat_formal_2}
\begin{aligned}
(\exists i, \forall j) (0 \leq i < |F|) \wedge (0 \leq j < |F[i]|) &\wedge ((isPos(F[i][j]) \wedge \neg \mathcal{A}[decode(F[i][j])]) \\
&\vee (\neg isPos(F[i][j]) \wedge \mathcal{A}[decode(F[i][j])])))
\end{aligned}
\end{gather*}

@@

%----------------------------------------------------------------------------------------
%	COPRIME
%----------------------------------------------------------------------------------------
\subsection{Two numbers are coprime}\label{subsec:coprime}
The greatest common divisor ($gcd$) of two non zero integers is the largest integer that divides both evenly \cite{knuthtaocp02}. A more specific $gcd$-problem is to find two numbers which are coprime, also called relative prime. Coprime expresses the relation between two integers a and b, where the greatest common divisor between those two is one \cite{coprimewiki}, i.e., $gcd(a, b) = 1$. To check if two numbers are coprime it has to be ensured that there is no greater $gcd$ than one, which can be stated in a propositional formula.
\begin{equation}\label{eq:coprime-universial}
    (\forall n) (2 \le n \le \min(a,b)) \Rightarrow \neg((a \mathbin{\%} n) = 0 \land (b \mathbin{\%} n) = 0)
\end{equation}
To check this it would take $\mathcal{O}(a)$, means linear time complexity. To find a contradiction to this condition we can formulate:
\begin{equation}\label{eq:coprime-existential}
    (\exists n) (2 \le n \le \min(a,b)) \land (a \mathbin{\%} n) = 0 \land (b \mathbin{\%} n) = 0
\end{equation}
When it is ensured that the greater number is not evenly divided by the smaller one ($\max(a,b)\mathbin{\%}\min(a,b)=0$), which automatically would then be a counterexample, the search space can reduced by half to $(2 \le n \le \lfloor \frac{a}{2} \rfloor)$, because in the upper half of the lower number no integer can evenly divide this number.

@@

\subsubsection{Least common multiple}
A similar problem would be to check the least common multiple ($lcm$) of two numbers. There the space between the greater number and the stated $lcm$ has to be checked for a smaller one. This search space can be significantly reduced by only selecting numbers, which are multiples of the greater number and check if the smaller number can divide this number without remainder.
\begin{equation}\label{eq:lcm-existential}
    (\exists n) (2 \le n < \frac{lcm}{\max(a,b)}) \land ((\max(a,b)*n) \mathbin{\%} \min(a,b) = 0)
\end{equation}

@@

\begin{lstlisting}[language=Solidity, style=Solidity, numbers=none, caption=Access a heap in array representation]
heap_array[(k-1)/2] // Get parent of actual node k
heap_array[(k*2)+1] // Get left child of actual node k
heap_array[(k*2)+2] // Get right child of actual node k
\end{lstlisting}
So to check if an array $a$ fulfills the min heap property, we can formulate it in the following predicate logic:
\begin{equation}\label{eq:heap-unversial_1}
  (\forall k) (1 \le k < |a|) \Rightarrow a[\lfloor(k-1)/2)\rfloor] \le a[k]
\end{equation}
This equation checks the direction from the child to the parent node. (Following the Exposé) The direction could also be turned around and every parent node could be checked against its child nodes, which does not change the number of comparisons.
\begin{equation}\label{eq:heap-unversial_2}
  (\forall k) (0 \le k \le \lfloor |a|/2 \rfloor) \Rightarrow a[k] \le a[2k+1] \land (2k+2 < |a| \Rightarrow a[k] \le a[2k+2])
\end{equation}
This test would need $\mathcal{O}(|a|)$, i.e., linear time complexity. To find a contradiction this can be formulated to
\begin{equation}\label{eq:heap-unversial_3}
  (\exists k) (1 \le k \le |a|) \land a[\lfloor(k-1)/2)\rfloor] > a[k]
\end{equation}
This example is similar to the example for checking if an array is sorted. Further such examples could be:
\begin{itemize}
    \item One array is the reverse of another
    \item Array fulfills binary search tree property (If not balanced, empty fields must be identified correctly)
\end{itemize}
